---
title: "Cetacean Distribution Analysis — Project Report"
author: "Laura Joan Feyrer"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: hide
    fig_width: 10
    fig_height: 6
---

```{r header-comment, include=FALSE}
# ============================================================
# Laura Joan Feyrer
# Date Updated: 2026-05-22
# Report: Cetacean_dist_report
# Description: RMarkdown report summarising cetacean
#              distribution analysis. Draws all figures from
#              saved PNGs on disk (no spatial data loaded);
#              summary statistics sourced from lightweight
#              CSV exports only. 

# ============================================================
```

```{r setup, include=FALSE}
# ---- Setup: global chunk options ----
knitr::opts_chunk$set(
  echo      = FALSE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center"
)

knitr::opts_knit$set(root.dir = here::here())
```

```{r packages, include=FALSE}
# ---- Setup: packages (lightweight only — no sf/raster/spatstat) ----
suppressMessages({
  if (!requireNamespace("pacman", quietly = TRUE)) {
    install.packages("pacman")
  }

  pacman::p_load(
    tidyverse,
    knitr,
    kableExtra,
    here,
    lubridate,
    scales,
    conflicted
  )
})

conflicted::conflict_prefer("filter", "dplyr")
```

```{r fig-paths, include=FALSE}
# ---- CSV freshness check --------------------------------------------------------
# Warns if confidence CSVs are older than the seasonal summary, which would
# indicate 05_summarize_wsdb_confidence.R was not re-run after the last data update.

check_csv_freshness <- function(reference_path, check_paths) {
  if (!file.exists(reference_path)) return(invisible())
  ref_time <- file.mtime(reference_path)
  stale <- check_paths[file.exists(check_paths) & 
                        file.mtime(check_paths) < ref_time]
  if (length(stale) > 0) {
    warning(
      "The following CSVs are older than seasonal_summary_WEA_vs_study.csv ",
      "and may contain stale counts. Re-run 04_summarize_wsdb_effort.R to update:\n  ",
      paste(basename(stale), collapse = "\n  ")
    )
  }
}

check_csv_freshness(
  reference_path = here("output/data/seasonal_summary_WEA_vs_study.csv"),
  check_paths    = c(
    here("output/data/overall_conf_summary.csv"),
    here("output/data/target_conf_summary.csv"),
    here("output/data/all_cet_seasonal_totals.csv")
  )
)
# ============================================================
# Figure & data path configuration
# All paths hardcoded from actual output/figs file listing
# Root: here() = project root (set by knitr::opts_knit above)
# ============================================================
# ---- Helper: restore apostrophes lost in filename conventions ----
# Called in label-cleaning functions before str_remove_all strips them.
# Covers the pattern where "Sowerby's" → "SOWERBYS" in filename →
# "Sowerbys" after toTitleCase → needs "Sowerby's" restored for display.
restore_apostrophes <- function(x) {
  x |>
    stringr::str_replace_all("(?i)\\bSowerby S\\b",    "Sowerby's") |>
    stringr::str_replace_all("(?i)\\bRisso S\\b",       "Risso's") |>
    stringr::str_replace_all("(?i)\\bTrue S\\b",        "True's") |>
    stringr::str_replace_all("(?i)\\bGervais S\\b",     "Gervais'") |>
    stringr::str_replace_all("(?i)\\bBlainville S\\b",  "Blainville's") |>
    stringr::str_replace_all("(?i)\\bCuvier S\\b",      "Cuvier's")
}
# ---- Helper: display a saved PNG or a fallback message ----
show_fig <- function(path) {
  if (!is.null(path) && file.exists(path)) {
    knitr::include_graphics(path)
  } else {
    cat(paste0("> *Figure not found: `", path, "`*\n\n"))
  }
}

# ---- Helper: loop-render a vector of PNG paths with knitr::include_graphics ----
# Use inside results='asis' chunks. knitr::include_graphics() is the only
# reliable way to embed images from R code — markdown ![]() syntax is not
# processed when emitted from cat() inside a chunk.
show_fig_loop <- function(files, label_fn) {
  for (f in files) {
    label <- label_fn(basename(f))
    cat(paste0("\n\n#### ", label, "\n\n"))
    # knitr::include_graphics needs a separate chunk call — we use
    # a sub-knit trick: emit a fenced chunk that will be evaluated
    # NOTE: for results='asis' loops the cleanest approach is to knit
    # child chunks. We use knitr::knit_expand + knitr::knit_child below.
  }
}

# ---- Helper: read CSV with graceful fallback ----
read_csv_safe <- function(path) {
  if (file.exists(path)) {
    readr::read_csv(path, show_col_types = FALSE)
  } else {
    cat(paste0("> *CSV not found: `", path,
               "` — add write_csv() to the relevant script and re-run once.*\n\n"))
    NULL
  }
}

# ==============================================================
# SIGHTINGS MAPS
# Source script: 03_plot_combined_sightings.R
# ggsave → output/figs/sights/
# ==============================================================
fig_wsdb_facet        <- here("output/figs/sights/WSDB_Whale_sightings_facet.png")
fig_aerial_facet      <- here("output/figs/sights/Aerial_Whale_sightings_facet.png")

# Individual WSDB species maps (output/figs/sights/WSDB_Sightings_*.png)
fig_wsdb_blue         <- here("output/figs/sights/WSDB_Sightings_WHALE_BLUE.png")
fig_wsdb_fin          <- here("output/figs/sights/WSDB_Sightings_WHALE_FIN.png")
fig_wsdb_humpback     <- here("output/figs/sights/WSDB_Sightings_WHALE_HUMPBACK.png")
fig_wsdb_minke        <- here("output/figs/sights/WSDB_Sightings_WHALE_MINKE.png")
fig_wsdb_sei          <- here("output/figs/sights/WSDB_Sightings_WHALE_NORTHERN_BOTTLENOSE.png")
fig_wsdb_narw         <- here("output/figs/sights/WSDB_Sightings_WHALE_NORTH_ATLANTIC_RIGHT.png")  # not in listing — check script
fig_wsdb_narw         <- here("output/figs/sights/WSDB_Sightings_WHALE_SPERM.png")                 # placeholder; update if NARW file exists
fig_wsdb_pilot        <- here("output/figs/sights/WSDB_Sightings_WHALE_LONG_FINNED_PILOT.png")
fig_wsdb_sperm        <- here("output/figs/sights/WSDB_Sightings_WHALE_SPERM.png")
fig_wsdb_sowerby      <- here("output/figs/sights/WSDB_Sightings_WHALE_SOWERBY_S_BEAKED.png")
fig_wsdb_dolphins_aws <- here("output/figs/sights/WSDB_Sightings_DOLPHINS_ATLANTIC_WHITE_SIDED.png")
fig_wsdb_dolphins_com <- here("output/figs/sights/WSDB_Sightings_DOLPHINS_COMMON.png")
fig_wsdb_dolphins_ris <- here("output/figs/sights/WSDB_Sightings_DOLPHINS_RISSO_S.png")
fig_wsdb_porpoise     <- here("output/figs/sights/WSDB_Sightings_PORPOISE_HARBOUR.png")            # not in listing — check
fig_wsdb_porpoise     <- here("output/figs/sights/WSDB_Sightings_DOLPHINS_PORPOISE_NS.png")        # closest match in listing

# ==============================================================
# figs/Effort_maps & COVERAGE MAPS
# Source script: 04_summarize_wsdb_effort.R
# ggsave → output/figs/figs/Effort_maps/
# ==============================================================
fig_coverage          <- here("output/figs/Effort_maps/00_all_cetacean_records.png")
  # p_all: gridded density of all cetacean sightings (25 km grid, log10 scale)
fig_confidence        <- here("output/figs/Effort_maps/00_data_consistency.png")
  # p_conf: coverage × temporal consistency confidence index
fig_seasonal_map      <- here("output/figs/Effort_maps/00_seasonal_coverage.png")
  # p_seasonal: seasonal faceted sightings map (Spring/Summer/Fall/Winter)
fig_seasonal_bars     <- here("output/figs/Effort_maps/00_seasonal_coverage_pct_study_area.png")
  # p_seasonal_bars: total records and occupied grid cells by season

# Target species figs/Effort_maps maps (records + figs/Effort_maps-normalised share)
Blue_whale_1_target_records       <- here("output/figs/Effort_maps/Blue_whale_1_target_records.png")
Blue_whale_2_share <- here("output/figs/Effort_maps/Blue_whale_2_share.png")
NARW_1_target_records       <- here("output/figs/Effort_maps/NARW_1_target_records.png")
NARW_2_share <- here("output/figs/Effort_maps/NARW_2_share.png")
Shelf_baleen_1_target_records      <- here("output/figs/Effort_maps/Shelf_baleen_1_target_records.png")
Shelf_baleen_2_share <- here("output/figs/Effort_maps/Shelf_baleen_2_share.png")
Deep_divers_1_target_records   <- here("output/figs/Effort_maps/Deep_divers_1_target_records.png")
Deep_divers_2_share<- here("output/figs/Effort_maps/Deep_divers_2_share.png")
Harbour_porpoise_1_target_records         <- here("output/figs/Effort_maps/Harbour_porpoise_1_target_records.png")
Harbour_porpoise_2_share   <- here("output/figs/Effort_maps/Harbour_porpoise_2_share.png")
Small_odontocetes_1_target_records   <- here("output/figs/Effort_maps/Small_odontocetes_1_target_records.png")
Small_odontocetes_2_share   <- here("output/figs/Effort_maps/Small_odontocetes_2_share.png")

# Confidence summary figures (05_summarize_wsdb_confidence.R)
fig_conf_comparison   <- here("output/figs/Effort_maps/confidence_summaries/confidence_comparison.png")
fig_conf_heatmap      <- here("output/figs/Effort_maps/confidence_summaries/target_confidence_heatmap.png")
fig_conf_heatmap_WEA  <- here("output/figs/Effort_maps/confidence_summaries/target_confidence_heatmap_OSW.png")

# ==============================================================
# PAM SUMMARY FIGURES
# Source script: 07_summarize_pam_detections.R
# ggsave → output/figs/figs/Effort_maps/PAM/
# ==============================================================
fig_pam_sites_year    <- here("output/figs/Effort_maps/PAM/annual_sitfigs/Effort_maps_bymonth.png")
  # p1: number of PAM sites sampled per year by season
fig_pam_combined      <- here("output/figs/Effort_maps/PAM/combined_seasonal.png")
  # combined_seasonal: p1–p5 patchwork layout saved to disk
fig_pam_monthly_det   <- here("output/figs/Effort_maps/PAM/monthly_detections_SS.png")
  # monthly detection summary
fig_pam_site_map      <- here("output/figs/Effort_maps/PAM/Site_effort_map.png")
  # p_map: PAM station distribution map by season
fig_pam_coverage      <- here("output/figs/Effort_maps/PAM/PAM_station_range_scenarios.png")
  # p_pam_range: 5/10/25/50 km acoustic buffer scenarios
fig_pam_station_cov   <- here("output/figs/Effort_maps/PAM/PAM_station_coverage.png")
  # overall PAM station coverage map

# ==============================================================
# KDE OUTPUTS — individual and family-level
# Source script: 08_run_kde_models.R → performKDE()
# ggsave → output/figs/
# Note: two versions per species —
#   KDE_*.png         = contour-only map (used in shapefiles/report)
#   KDE_*.png     = full annotated figure with basemap
# ==============================================================

# ---- PAM KDEs: baleen whales ----
fig_kde_pam_blue      <- here("output/figs/KDE_maps/KDE_pam_baleen_Blue_Whale_Bm_fixed_bw10000.png")
fig_kde_pam_fin       <- here("output/figs/KDE_maps/KDE_pam_baleen_Fin_Whale_Bp_fixed_bw10000.png")
fig_kde_pam_humpback  <- here("output/figs/KDE_maps/KDE_pam_baleen_Humpback_Whale_Mn_fixed_bw10000.png")
fig_kde_pam_minke     <- here("output/figs/KDE_maps/KDE_pam_baleen_Minke_Whale_Ba_fixed_bw10000.png")
fig_kde_pam_narw      <- here("output/figs/KDE_maps/KDE_pam_baleen_North_Atlantic_Right_Whale_Eg_fixed_bw10000.png")
fig_kde_pam_sei       <- here("output/figs/KDE_maps/KDE_pam_baleen_Sei_Whale_Bb_fixed_bw10000.png")

# ---- PAM KDEs: beaked whales ----
fig_kde_pam_zc        <- here("output/figs/KDE_maps/KDE_pam_beaked_Goose_Beaked_Whale_Zc_fixed_bw10000.png")
fig_kde_pam_ha        <- here("output/figs/KDE_maps/KDE_pam_beaked_Northern_Bottlenose_Whale_Ha_fixed_bw10000.png")
fig_kde_pam_mb        <- here("output/figs/KDE_maps/KDE_pam_beaked_Sowerbys_Beaked_Whale_Mb_fixed_bw10000.png")
fig_kde_pam_mmme      <- here("output/figs/KDE_maps/KDE_pam_beaked_Trues_Gervais_Beaked_Whale_MmMe_fixed_bw10000.png")

# ---- Sightings KDEs: family groups ----
fig_kde_sights_mysticetes  <- here("output/figs/KDE_maps/KDE_sightings_mysticetes_Mysticetes_(All_Baleen_Whales)_fixed_bw10000.png")
fig_kde_sights_odontocetes <- here("output/figs/KDE_maps/KDE_sightings_odontocetes_Odontocetes_(All_Toothed_Whales)_fixed_bw10000.png")

# ---- Sightings KDEs: individual species (proximity-weighted) ----
fig_kde_sights_blue        <- here("output/figs/KDE_maps/KDE_sightings_species_proximity_WHALE-BLUE_fixed_bw10000.png")
fig_kde_sights_fin         <- here("output/figs/KDE_maps/KDE_sightings_species_proximity_WHALE-FIN_fixed_bw10000.png")
fig_kde_sights_humpback    <- here("output/figs/KDE_maps/KDE_sightings_species_proximity_WHALE-HUMPBACK_fixed_bw10000.png")
fig_kde_sights_minke       <- here("output/figs/KDE_maps/KDE_sightings_species_proximity_WHALE-MINKE_fixed_bw10000.png")
fig_kde_sights_narw        <- here("output/figs/KDE_maps/KDE_sightings_species_proximity_WHALE-NORTH_ATLANTIC_RIGHT_fixed_bw10000.png")
fig_kde_sights_sei         <- here("output/figs/KDE_maps/KDE_sightings_species_proximity_WHALE-SEI_fixed_bw10000.png")
fig_kde_sights_bottlenose  <- here("output/figs/KDE_maps/KDE_sightings_species_proximity_WHALE-NORTHERN_BOTTLENOSE_fixed_bw10000.png")
fig_kde_sights_sperm       <- here("output/figs/KDE_maps/KDE_sightings_species_proximity_WHALE-SPERM_fixed_bw10000.png")
fig_kde_sights_pilot       <- here("output/figs/KDE_maps/KDE_sightings_species_proximity_WHALE-_LONG-FINNED_PILOT_fixed_bw10000.png")
fig_kde_sights_sowerby     <- here("output/figs/KDE_maps/KDE_sightings_species_proximity_WHALE-SOWERBY'S_BEAKED_fixed_bw10000.png")
fig_kde_sights_aws         <- here("output/figs/KDE_maps/KDE_sightings_species_proximity_DOLPHINS-ATLANTIC_WHITE-SIDED_fixed_bw10000.png")
fig_kde_sights_common_dol  <- here("output/figs/KDE_maps/KDE_sightings_species_proximity_DOLPHINS-COMMON_fixed_bw10000.png")
fig_kde_sights_risso       <- here("output/figs/KDE_maps/KDE_sightings_species_proximity_DOLPHINS-RISSO'S_fixed_bw10000.png")
fig_kde_sights_harbour_p   <- here("output/figs/KDE_maps/KDE_sightings_species_proximity_PORPOISE-HARBOUR_fixed_bw10000.png")
fig_kde_sights_baleen_ns   <- here("output/figs/KDE_maps/KDE_sightings_species_proximity_BALEEN_WHALE_(NS)_fixed_bw10000.png")
fig_kde_sights_dol_p_ns    <- here("output/figs/KDE_maps/KDE_sightings_species_proximity_DOLPHINS_PORPOISE_(NS)_fixed_bw10000.png")

# ---- Grouped KDEs ----
fig_kde_grouped_baleen     <- here("output/figs/KDE_maps/KDE_pam_grouped_baleen_All_Baleen_Whales_(PAM)_fixed_bw10000.png")
fig_kde_grouped_beaked     <- here("output/figs/KDE_maps/KDE_pam_grouped_beaked_All_Beaked_Whales_(PAM)_fixed_bw10000.png")

# ==============================================================
# PAM vs SIGHTINGS COMPARISON MAPS
# Source script: 10_compare_kde_pam_vs_sightings.R → create_comparison_map()
# ggsave → output/figs/compare_sightings_pam/
# Naming: compare_<species>__S_<method>__P_<method>.png
#         pam_only_<species>__P_<method>.png
#         sightings_only_<species>__S_<method>.png
# ==============================================================

# ---- Side-by-side PAM vs sightings ----
fig_cmp_blue          <- here("output/figs/compare_sightings_pam/compare_blue_whale__S_prox_fixed_bw10000__P_unw_fixed_bw10000.png")
fig_cmp_fin           <- here("output/figs/compare_sightings_pam/compare_fin_whale__S_prox_fixed_bw10000__P_unw_fixed_bw10000.png")
fig_cmp_humpback      <- here("output/figs/compare_sightings_pam/compare_humpback_whale__S_prox_fixed_bw10000__P_unw_fixed_bw10000.png")
fig_cmp_minke         <- here("output/figs/compare_sightings_pam/compare_minke_whale__S_prox_fixed_bw10000__P_unw_fixed_bw10000.png")
fig_cmp_narw          <- here("output/figs/compare_sightings_pam/compare_north_atlantic_right_whale__S_prox_fixed_bw10000__P_unw_fixed_bw10000.png")
fig_cmp_bottlenose    <- here("output/figs/compare_sightings_pam/compare_northern_bottlenose_whale__S_prox_fixed_bw10000__P_unw_fixed_bw10000.png")
fig_cmp_sei           <- here("output/figs/compare_sightings_pam/compare_sei_whale__S_prox_fixed_bw10000__P_unw_fixed_bw10000.png")
fig_cmp_sowerby       <- here("output/figs/compare_sightings_pam/compare_sowerbys_beaked_whale__S_prox_fixed_bw10000__P_unw_fixed_bw10000.png")
fig_cmp_all_baleen    <- here("output/figs/compare_sightings_pam/compare_all_baleen_whales__S_unw_fixed_bw10000__P_unw_fixed_bw10000.png")

# ---- PAM only (no sightings KDE available) ----
fig_pam_only_beaked   <- here("output/figs/compare_sightings_pam/pam_only_all_beaked_whales__P_unw_fixed_bw10000.png")
fig_pam_only_zc       <- here("output/figs/compare_sightings_pam/pam_only_goose_beaked_whale__P_unw_fixed_bw10000.png")
fig_pam_only_mmme     <- here("output/figs/compare_sightings_pam/pam_only_trues_gervais_beaked_whale__P_unw_fixed_bw10000.png")

# ---- Sightings only (no PAM KDE available) ----
fig_so_odontocetes    <- here("output/figs/compare_sightings_pam/sightings_only_all_odontocetes__S_unw_fixed_bw10000.png")
fig_so_baleen_ns      <- here("output/figs/compare_sightings_pam/sightings_only_baleen_whale_ns___S_prox_fixed_bw10000.png")
fig_so_aws            <- here("output/figs/compare_sightings_pam/sightings_only_dolphins_atlantic_white_sided__S_prox_fixed_bw10000.png")
fig_so_common_dol     <- here("output/figs/compare_sightings_pam/sightings_only_dolphins_common__S_prox_fixed_bw10000.png")
fig_so_risso          <- here("output/figs/compare_sightings_pam/sightings_only_dolphins_risso_s__S_prox_fixed_bw10000.png")
fig_so_harbour_p      <- here("output/figs/compare_sightings_pam/sightings_only_porpoise_harbour__S_prox_fixed_bw10000.png")
fig_so_pilot          <- here("output/figs/compare_sightings_pam/sightings_only_whale_long_finned_pilot__S_prox_fixed_bw10000.png")
fig_so_sperm          <- here("output/figs/compare_sightings_pam/sightings_only_whale_sperm__S_prox_fixed_bw10000.png")

# ==============================================================
# STUDY AREA MAP
# Source script: setup/02_build_study_area.R / helper_spatial.R
# ==============================================================
fig_study_area        <- here("output/figs/Effort_maps/00_all_cetacean_records.png")

# ==============================================================
# CSV SUMMARY EXPORTS
# Add write_csv() calls to each source script (see chunk comments below)
# ==============================================================
csv_sightings_summary    <- here("output", "data", "sightings_summary.csv")
csv_sightings_by_source  <- here("output", "data", "sightings_by_source.csv")
csv_sightings_by_species <- here("output", "data", "sightings_by_species.csv")
csv_pam_summary          <- here("output", "data", "pam_summary.csv")
csv_pam_seasonal         <- here("output", "data", "pam_seasonal_summary.csv")
csv_conf_overall         <- here("output", "data", "overall_conf_summary.csv")
csv_conf_target          <- here("output", "data", "target_conf_summary.csv")
csv_appendix_a_table1    <- here("output", "data", "appendix_a_table1_wea_documented_species.csv")
csv_appendix_a_table2    <- here("output", "data", "appendix_a_table2_wea_comparability.csv")
csv_appendix_a_table2_stats <- here("output", "data", "appendix_a_table2_wea_comparability_stats.csv")
csv_appendix_a_species_status <- here("output", "data", "appendix_a_wea_species_status.csv")
csv_wea_10km_species    <- here("output", "data", "wea_10km_species_occurrence_summary.csv")
csv_wea_10km_area       <- here("output", "data", "wea_10km_area_summary.csv")
csv_pam_wea_distance    <- here("output", "data", "pam_wea_distance_summary.csv")
thresholds <- readRDS(here("output/data/sightings_thresholds.rds"))

```

---

# Figures & Plots {#figures}

## Study Area {#study-area}

```{r fig-study-area, fig.cap="Map of study area"}
# output/figs/proposed_StudyAreamap.png
# Source: setup/02_build_study_area.R / helper_spatial.R
show_fig(fig_study_area)
```

---

## Cetacean Sightings Distribution Maps 2010 - 2024{#sightings-maps}

### WSDB — Opportunistic Sightings
Species were retained for mapping if they had at least 
`r thresholds$wsdb` records in the WSDB dataset
```{r fig-wsdb-facet,  fig.height=8}
# output/figs/sights/WSDB_Whale_sightings_facet.png
# Source: 03_plot_combined_sightings.R → Map_Species facet ggsave

show_fig(fig_wsdb_facet)
```

### Aerial Survey — Systematic Sightings
Species were retained for mapping if they had at least 
`r thresholds$aerial` records in the aerial survey dataset

```{r fig-aerial-facet,  fig.height=8}
# output/figs/sights/Aerial_Whale_sightings_facet.png
# Source: 03_plot_combined_sightings.R → aerial facet ggsave
show_fig(fig_aerial_facet)
```

### Individual Species Maps 2010-2024
Species shown based on minimum record thresholds

```{r fig-sightings-species, results='asis'}
# output/figs/sights/WSDB_Sightings_*.png and Aerial_Sightings_*.png
# Auto-renders all individual species maps present on disk.
# Source: 03_plot_combined_sightings.R -> per-species loop -> ggsave

sights_dir <- here("output/figs/sights")

clean_sights_label <- function(fname) {
  fname |>
    tools::file_path_sans_ext() |>
    stringr::str_replace("^(WSDB|Aerial)_Sightings_", "\\1 - ") |>
    stringr::str_replace_all("_", " ") |>
    stringr::str_to_title() |>
restore_apostrophes() |>
    stringr::str_remove_all("'")  # restore first, then strip for safe chunk labels
  }

sights_species_files <- list.files(
  sights_dir,
  pattern    = "^(WSDB|Aerial)_Sightings_.*\\.png$",
  full.names = TRUE
)

# Sort NS species to end within each source, matching facet plot order
ns_last_files <- function(paths) {
  c(sort(paths[!grepl("_NS_|_Ns_|(NS)", paths)]),
    sort(paths[grepl("_NS_|_Ns_|(NS)", paths)]))
}

# Split by source, reorder each, recombine
wsdb_files   <- ns_last_files(sights_species_files[grepl("^WSDB", basename(sights_species_files))])
aerial_files <- ns_last_files(sights_species_files[grepl("^Aerial", basename(sights_species_files))])
sights_species_files <- c(wsdb_files, aerial_files)

if (length(sights_species_files) > 0) {
  chunks <- lapply(sights_species_files, function(f) {
    label  <- clean_sights_label(basename(f))
    f_safe <- gsub("'", "\\'", f, fixed = TRUE)  # escape apostrophes in path
    knitr::knit_expand(
      text = paste0(
        "\n\n#### {{label}}\n\n",
        strrep("`", 3), "{r echo=FALSE}\n",
        "knitr::include_graphics(\"{{f_safe}}\")\n",
        strrep("`", 3), "\n"
      ),
      label = label, f_safe = f_safe
    )
  })
  cat(knitr::knit_child(text = unlist(chunks), quiet = TRUE), sep = "\n")
} else {
  cat("> *No individual species sightings maps found in `output/figs/sights/`.*\n\n")
}
```

---

## Effort & Coverage Maps {#Effort-maps}

### All Cetacean Records — Coverage Density

```{r fig-coverage, fig.cap="Gridded density of all cetacean sightings (25 km grid, log10 scale). OWA boundaries in orange; study area dashed."}
# output/figs/figs/Effort_maps/00_all_cetacean_records.png
# Source: 04_summarize_wsdb_effort.R → p_all
show_fig(fig_coverage)
```

### Data Confidence Map

```{r fig-confidence, fig.cap="Coverage × temporal consistency confidence index. Green = high confidence."}
# output/figs/figs/Effort_maps/00_data_consistency.png
# Source: 04_summarize_wsdb_effort.R → p_conf
show_fig(fig_confidence)
```

### Seasonal Sightings — Faceted Map

```{r fig-seasonal-map, fig.cap="Seasonal distribution of all cetacean sightings on 25 km grid (log10 scale). Spring Mar–May, Summer Jun–Aug, Fall Sep–Nov, Winter Dec–Feb.", fig.height=9}
# output/figs/figs/Effort_maps/00_seasonal_coverage.png
# Source: 06_summarize_wsdb_seasonality.R → p_seasonal
show_fig(fig_seasonal_map)
```

### Seasonal Sightings — Coverage Summary

```{r fig-seasonal-bars, fig.cap=""}
# output/figs/figs/Effort_maps/00_seasonal_comparison_records_per_km2.png
# Source: 06_summarize_wsdb_seasonality.R → p_seasonal_bars
show_fig(fig_seasonal_bars)
```

---

## Target Species Maps {#target-maps}

### Blue Whale

```{r blue-whale-effort-maps,fig.cap="Blue whale — target records (left) and effort-normalised share (right).", fig.show='hold', out.width='49%',fig.align='center'}

knitr::include_graphics(c(
  Blue_whale_1_target_records,
  Blue_whale_2_share
))
```

### North Atlantic Right Whale

```{r NARW-effort-maps, fig.cap="NARW — target records (left) and Effort-normalised share (right).", fig.show='hold', out.width='49%',fig.align='center'}
# output/figs/figs/Effort_maps/NARW_1_target_records.png + _2_share.png

knitr::include_graphics(c(NARW_1_target_records,
NARW_2_share))
```

### Shelf Baleen

```{r baleen_maps, fig.cap="Shelf baleen whales — target records and Effort-normalised share.", fig.show='hold', out.width='49%',fig.align='center'}
# output/figs/figs/Effort_maps/Shelf_baleen_1_target_records.png + _2_share.png
knitr::include_graphics(c(Shelf_baleen_1_target_records,Shelf_baleen_2_share))
```

### Deep Divers (Beaked Whales)

```{r beaked_maps, fig.cap="Deep divers — target records and Effort-normalised share.", fig.show='hold', out.width='49%',fig.align='center'}
# output/figs/figs/Effort_maps/Deep_divers_1_target_records.png + _2_share.png
knitr::include_graphics(c(Deep_divers_1_target_records,
Deep_divers_2_share))
```

### Harbour Porpoise

```{r HP_maps, fig.cap="Harbour porpoise — target records and Effort-normalised share.", fig.show='hold', out.width='49%',fig.align='center'}
# output/figs/figs/Effort_maps/Harbour_porpoise_1_target_records.png + _2_share.png
knitr::include_graphics(c(Harbour_porpoise_1_target_records,
Harbour_porpoise_2_share))
```

### Small Odontocetes

```{r sm_odontocete_maps, fig.cap="Small odontocetes — target records and Effort-normalised share.", fig.show='hold', out.width='49%',fig.align='center'}
# output/figs/figs/Effort_maps/Small_odontocetes_1_target_records.png + _2_share.png
knitr::include_graphics(c(Small_odontocetes_1_target_records,
Small_odontocetes_2_share))
```

---

## Data Confidence Summaries {#confidence-figs}

```{r fig-conf-comparison, fig.cap="Confidence comparison across study region vs WEA areas."}
# output/figs/Effort_maps/confidence_summaries/confidence_comparison.png
# Source: 05_summarize_wsdb_confidence.R
show_fig(fig_conf_comparison)
```

```{r fig-conf-heatmap-WEA, fig.cap="Target species confidence heatmap — WEA areas only."}
# output/figs/Effort_maps/confidence_summaries/target_confidence_heatmap_WEA.png
show_fig(fig_conf_heatmap_WEA)
```

---

## PAM Summaries {#pam-summaries}

### Combined Seasonal Summary 

```{r fig-pam-combined, fig.cap="PAM seasonal summary: detection proportions, detection days", fig.height=10}
# output/figs/Effort_maps/PAM/combined_seasonal.png
# Source: 07_summarize_pam_detections.R → combined_seasonal patchwork → ggsave
show_fig(fig_pam_combined)
```

### PAM Station Distribution Map

```{r fig-pam-site-map, fig.cap="PAM station locations by season. Point transparency = years sampled. OWA boundaries in orange.", fig.height=9}
# output/figs/Effort_maps/PAM/Site_figs/Effort_maps_map.png
# Source: 07_summarize_pam_detections.R → p_map
show_fig(fig_pam_site_map)
```

### PAM Station Coverage Map

```{r fig-pam-station-cov, fig.cap="Overall PAM station spatial coverage."}
# output/figs/Effort_maps/PAM/PAM_station_coverage.png
# Source: 07_summarize_pam_detections.R → p_pam_range (coverage version)
show_fig(fig_pam_station_cov)
```

---

## KDE Outputs {#kde-outputs}

> All KDE figures from `08_run_kde_models.R → performKDE()`


### Sightings KDEs — Family Groups

```{r fig-kde-mysticetes, fig.cap="Sightings: Mysticetes (all baleen whales)"}
# output/figs/KDE_maps/KDE_sightings_mysticetes_Mysticetes_(All_Baleen_Whales)_fixed_bw10000.png
show_fig(fig_kde_sights_mysticetes)
```

```{r fig-kde-odontocetes, fig.cap="Sightings: Odontocetes (all toothed whales)"}
# output/figs/KDE_maps/KDE_sightings_odontocetes_Odontocetes_(All_Toothed_Whales)_fixed_bw10000.png
show_fig(fig_kde_sights_odontocetes)
```

### Sightings KDEs — Individual Species (proximity-weighted)
Species shown reflect whatever passed `min_records_species` in `08_run_kde_models.R` at last run — threshold changes are automatically reflected.

```{r fig-kde-sights-species, results='asis'}
# output/figs/KDE_maps/KDE_sightings_species_proximity_*.png
# Auto-renders all individual sightings KDEs on disk.
# Source: 08_run_kde_models.R → performKDE() → ggsave

kde_sights_files <- list.files(
  here("output/figs/"),
  pattern    = "^KDE_sightings_species_proximity_.*\\.png$",
  full.names = TRUE
)

clean_kde_label <- function(fname) {
  fname |>
    tools::file_path_sans_ext() |>
    stringr::str_remove("^KDE_sightings_species_proximity_") |>
    stringr::str_remove("_fixed_bw\\d+$") |>
    stringr::str_replace_all("_", " ") |>
    stringr::str_to_title() |>
restore_apostrophes() |>
    stringr::str_remove_all("'")  # restore first, then strip for safe chunk labels
  }

if (length(kde_sights_files) > 0) {
  chunks <- lapply(kde_sights_files, function(f) {
    label  <- clean_kde_label(basename(f))
    f_safe <- gsub("'", "\\'", f, fixed = TRUE)  # escape apostrophes in path
    knitr::knit_expand(
      text = c(
        "\n\n#### {{label}}\n\n",
        "```{r echo=FALSE}",
        "knitr::include_graphics(\"{{f_safe}}\")",
        "```\n"
      ),
      label = label, f_safe = f_safe
    )
  })
  cat(knitr::knit_child(text = unlist(chunks), quiet = TRUE), sep = "\n")
} else {
  cat("> *No individual sightings KDE PNGs found in `output/figs/`.*\n\n")
}
```

### PAM KDEs — Baleen Whales 
- weighted by proportion of detection days

```{r fig-kde-pam-blue, fig.cap="PAM Blue whale (Bm)"}
# output/figs/KDE_maps/KDE_pam_baleen_Blue_Whale_Bm_fixed_bw10000.png
show_fig(fig_kde_pam_blue)
```

```{r fig-kde-pam-fin, fig.cap="PAM Fin whale (Bp) "}
show_fig(fig_kde_pam_fin)
```

```{r fig-kde-pam-humpback, fig.cap="PAM Humpback whale (Mn) "}
show_fig(fig_kde_pam_humpback)
```

```{r fig-kde-pam-minke, fig.cap="PAM Minke whale (Ba) "}
show_fig(fig_kde_pam_minke)
```

```{r fig-kde-pam-narw, fig.cap="PAM North Atlantic right whale (Eg) "}
show_fig(fig_kde_pam_narw)
```

```{r fig-kde-pam-sei, fig.cap="PAM: Sei whale (Bb)"}
show_fig(fig_kde_pam_sei)
```

### PAM KDEs — Beaked Whales
- weighted by proportion of detection days


```{r fig-kde-pam-zc, fig.cap="PAM Goose beaked whale (Zc) "}
show_fig(fig_kde_pam_zc)
```

```{r fig-kde-pam-ha, fig.cap="PAM Northern bottlenose whale (Ha) "}
show_fig(fig_kde_pam_ha)
```

```{r fig-kde-pam-mb, fig.cap="PAM Sowerby's beaked whale (Mb) "}
show_fig(fig_kde_pam_mb)
```

```{r fig-kde-pam-mmme, fig.cap="PAM True's / Gervais' beaked whale (MmMe) "}
show_fig(fig_kde_pam_mmme)
```

### Grouped PAM KDEs

```{r fig-kde-grouped-baleen, fig.cap="PAM All baleen whales grouped "}
# output/figs/KDE_maps/KDE_pam_grouped_baleen_All_Baleen_Whales_(PAM)_fixed_bw10000.png
show_fig(fig_kde_grouped_baleen)
```

```{r fig-kde-grouped-beaked, fig.cap="PAM All beaked whales grouped "}
# output/figs/KDE_maps/KDE_pam_grouped_beaked_All_Beaked_Whales_(PAM)_fixed_bw10000.png
show_fig(fig_kde_grouped_beaked)
```

---

## PAM vs Sightings Comparison Maps {#comparison-maps}

> Source: `10_compare_kde_pam_vs_sightings.R` → `output/figs/compare_sightings_pam/`
> Filename convention: `compare_<species>__S_<method>__P_<method>.png` (side-by-side),
> `pam_only_<species>__P_<method>.png`, `sightings_only_<species>__S_<method>.png`.
> All files present on disk are rendered — species added/removed in the script are automatically reflected.

```{r fig-compare-all, results='asis'}
# output/figs/compare_sightings_pam/*.png
# Auto-renders all comparison figures on disk, grouped by type.
# Source: 10_compare_kde_pam_vs_sightings.R → create_comparison_map() → ggsave

cmp_dir <- here("output/figs/compare_sightings_pam")

clean_cmp_label <- function(fname) {
  fname |>
    tools::file_path_sans_ext() |>
    stringr::str_remove("__(S|P)_.*$") |>   # strip method suffix
    stringr::str_replace_all("_", " ") |>
    stringr::str_to_title() |>
    stringr::str_remove_all("'")  # remove apostrophes — break chunk header parsing
}

cmp_files <- list.files(cmp_dir, pattern = "\\.png$", full.names = TRUE)

# Group by prefix type for clear section headers
compare_files   <- cmp_files[stringr::str_detect(basename(cmp_files), "^compare_")]
pam_only_files  <- cmp_files[stringr::str_detect(basename(cmp_files), "^pam_only_")]
sights_only_files <- cmp_files[stringr::str_detect(basename(cmp_files), "^sightings_only_")]

render_group <- function(files, heading) {
  if (length(files) == 0) return(invisible())
  cat(paste0("\n\n### ", heading, "\n\n"))
  chunks <- lapply(files, function(f) {
    label  <- clean_cmp_label(basename(f))
    f_safe <- gsub("'", "\\'", f, fixed = TRUE)  # escape apostrophes in path
    knitr::knit_expand(
      text = c(
        "\n\n#### {{label}}\n\n",
        "```{r echo=FALSE}",
        "knitr::include_graphics(\"{{f_safe}}\")",
        "```\n"
      ),
      label = label, f_safe = f_safe
    )
  })
  cat(knitr::knit_child(text = unlist(chunks), quiet = TRUE), sep = "\n")
}

if (length(cmp_files) == 0) {
  cat("> *No comparison PNGs found in `output/figs/compare_sightings_pam/`. Run `10_compare_kde_pam_vs_sightings.R`.*\n\n")
} else {
  render_group(compare_files,     "Side-by-side PAM vs Sightings")
  render_group(pam_only_files,    "PAM Only (no sightings KDE available)")
  render_group(sights_only_files, "Sightings Only (no PAM KDE available)")
}
```

---

# Key Parameters & Analytical Decisions {#parameters}

```{r params-table}
# ---- Key Parameters ----
# Read live from CSVs exported by each source script.
# If a file is missing, re-run the relevant script to regenerate it.

param_files <- c(
  here("output/data/params_kde.csv"),
  here("output/data/params_figs/Effort_maps.csv"),
  here("output/data/params_dedup.csv"),
  here("output/data/params_compare.csv")
)

params_df <- purrr::map(param_files, \(f) {
  if (file.exists(f)) {
    readr::read_csv(f, show_col_types = FALSE, col_types = readr::cols(.default = "c"))
  } else {
    message("Missing params file — re-run source script: ", basename(f))
    NULL
  }
}) |>
  purrr::compact() |>
  dplyr::bind_rows()

knitr::kable(
  params_df,
  caption   = "Key parameters and analytical decisions — live from source scripts",
  col.names = c("Parameter", "Value", "Script", "Notes"),
  align     = c("l", "r", "l", "l")
) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = TRUE,
    font_size  = 13
  ) %>%
  kableExtra::column_spec(1, bold = TRUE, width = "16em") %>%
  kableExtra::column_spec(2, width = "8em") %>%
  kableExtra::column_spec(3, italic = TRUE, width = "16em")
```

---

# Summary Statistics {#summary-stats}

> Summary stats are read from lightweight CSV exports.
> If a CSV is missing, add a `write_csv()` call to the relevant script and re-run it once.
> The path expected for each file is shown in the fallback message.

## Sightings Data {#sightings-stats}

```{r summary-sightings}
# ---- Summary: sightings dataset ----
# Expected file: output/data/sightings_summary.csv

df_sights <- read_csv_safe(csv_sightings_summary)
if (!is.null(df_sights)) {
  knitr::kable(df_sights, caption = "Sightings dataset overview") %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                              full_width = FALSE)
}
```

### Records by Source

```{r summary-source-table}
# ---- Summary: records by source ----
# Expected file: output/data/sightings_by_source.csv

df_source <- read_csv_safe(csv_sightings_by_source)
if (!is.null(df_source)) {
  knitr::kable(df_source,
             caption   = "Records by data source (post-deduplication, clipped to study area)",
             col.names = c("Source", "N Records", "N Species Codes", "N Species (NS)", "Date Range")) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
}
```

## Appendix C. Number of sightings records for species or groups in the study area

```{r summary-species-counts}
# ---- Appendix C: records by species with scientific names ----
# Expected file: output/data/sightings_by_species.csv

sci_lookup <- tibble::tribble(
  ~common_name,                    ~scientific_name,
  "Dolphin/Porpoise (NS)",         "—",
  "Cetacean (NS)",                 "—",
  "Fin/Sei Whale",                 "—",
  "Long-Finned Pilot Whale",       "Globicephala melas",
  "Northern Bottlenose Whale",     "Hyperoodon ampullatus",
  "Humpback Whale",                "Megaptera novaeangliae",
  "Common Dolphin",                "Delphinus delphis",
  "Minke Whale",                   "Balaenoptera acutorostrata",
  "Fin Whale",                     "Balaenoptera physalus",
  "Atlantic White-Sided Dolphin",  "Lagenorhynchus acutus",
  "Sowerby's Beaked Whale",        "Mesoplodon bidens",
  "Sperm Whale",                   "Physeter macrocephalus",
  "Blue Whale",                    "Balaenoptera musculus",
  "Risso's Dolphin",               "Grampus griseus",
  "Harbour Porpoise",              "Phocoena phocoena",
  "Atlantic Bottlenose Dolphin",   "Tursiops truncatus",
  "North Atlantic Right Whale",    "Eubalaena glacialis",
  "Striped Dolphin",               "Stenella coeruleoalba",
  "White-Beaked Dolphin",          "Lagenorhynchus albirostris",
  "Sei Whale",                     "Balaenoptera borealis",
  "Killer Whale",                  "Orcinus orca",
  "Cuvier's Beaked Whale",         "Ziphius cavirostris",
  "Beluga Whale",                  "Delphinapterus leucas",
  "False Killer Whale",            "Pseudorca crassidens"
)

df_sp <- read_csv_safe(csv_sightings_by_species)
if (!is.null(df_sp)) {
  df_sp %>%
    dplyr::arrange(dplyr::desc(n)) %>%
    dplyr::mutate(Rank = dplyr::row_number()) %>%
    dplyr::left_join(sci_lookup, by = "common_name") %>%
    dplyr::mutate(scientific_name = tidyr::replace_na(scientific_name, "—")) %>%
    dplyr::select(Rank,
                  `Common Name`      = common_name,
                  `Scientific Name`  = scientific_name,
                  `Sightings Records` = n) %>%
    knitr::kable(align = c("r","l","l","r"),
                 format.args = list(big.mark = ",")) %>%
    kableExtra::kable_styling(bootstrap_options = c("striped","hover","condensed"),
                              full_width = FALSE) %>%
    kableExtra::column_spec(3, italic = TRUE)
}
```

---

## PAM Data {#pam-stats}

```{r summary-pam}
# ---- Summary: PAM detection summary ----
# Expected file: output/data/pam_summary.csv


df_pam <- read_csv_safe(csv_pam_summary)
if (!is.null(df_pam)) {
  knitr::kable(df_pam,
               caption   = "PAM detection summary by species (all sites, all seasons)",
               col.names = c("Species Code", "N Sites", "Total Days Sampled",
                             "Detection Days", "Detection Rate")) %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                              full_width = FALSE)
}
```

### Seasonal PAM Summary (Study Area Stations)

```{r pam-seasonal-summary}
# ---- Summary: seasonal PAM coverage ----
# Expected file: output/data/pam_seasonal_summary.csv


df_pam_season <- read_csv_safe(csv_pam_seasonal)
if (!is.null(df_pam_season)) {
  knitr::kable(df_pam_season,
               caption   = "Seasonal coverage summary for study-area PAM stations",
               col.names = c("Season", "N Stations", "N Detection Days", "N Years")) %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                              full_width = FALSE)
}
```

---

## Data Confidence Summary {#confidence-stats}

```{r conf-summary}
# ---- Summary: overall confidence ----
# Expected file: output/data/overall_conf_summary.csv

df_conf <- read_csv_safe(csv_conf_overall)
if (!is.null(df_conf)) {
  knitr::kable(df_conf,
               caption = "Overall data confidence: study region vs WEAs",
               digits  = 1) %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
}
```

### Target Species Confidence (WEA Areas)

```{r target-conf-summary}
# ---- Summary: target species confidence ----
# Expected file: output/data/target_conf_summary.csv


df_conf_target <- read_csv_safe(csv_conf_target)
if (!is.null(df_conf_target)) {
  knitr::kable(df_conf_target,
               caption = "Target species/group confidence within WEAs",
               digits  = 1) %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
}
```

###Seasonal summary
```{r seasonal-setup, include=FALSE}
# ==============================================================================
# Laura Joan Feyrer
# Date Updated: 2026-03-12
# Chunk: seasonal_confidence_chunk.Rmd
# Description: Reads the seasonal summary CSV produced by 04_summarize_wsdb_effort.R and
#              renders (1) a formatted kable table of mean records/year by
#              season x species group x region, and (2) a bar chart showing
#              each species group's seasonal records as a percentage of total
#              all-cetacean records in that region and season. Deep divers
#              excluded (0 WEA records). Requires:
#                output/data/seasonal_summary_WEA_vs_study.csv
#                dt object in session (from 04_summarize_wsdb_effort.R) for seasonal totals
# Changes from previous version:
#   - Table changed from total records to mean records/year per species group
#     per season -- more comparable across groups of different abundance.
#   - Plot changed from raw counts to percentage of total cetacean records in
#     that region x season -- controls for summer dominance in figs/Effort_maps coverage.
#   - Removed cross-species seasonal average table (not sufficiently informative).
# ==============================================================================

library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(readr)
library(here)
conflicted::conflicts_prefer(
  dplyr::filter,
  dplyr::select,
  dplyr::mutate,
  dplyr::lag
)


# ---- Load seasonal summary ------------------------------------------------------

seasonal_wide <- readr::read_csv(
  here("output", "data", "seasonal_summary_WEA_vs_study.csv"),
  show_col_types = FALSE
)
# ---- Load all-cetacean seasonal totals (written by 04_summarize_wsdb_effort.R Section 15) ----
# These are raw totals across all species before any guild grouping, so there
# is no double-counting from guild rows inflating the denominator.
all_cet_seasonal <- readr::read_csv(
  here("output", "data", "all_cet_seasonal_totals.csv"),
  show_col_types = FALSE
) %>%
  mutate(
    season = factor(season, levels = c("Spring", "Summer", "Fall", "Winter")),
    region = factor(region, levels = c("Study Region", "WEA"))
  )

# ---- Pivot species summary to long and calculate metrics ------------------------

seasonal_long <- seasonal_wide %>%
  tidyr::pivot_longer(
    cols          = -c(target_label, region),
    names_to      = c("season", ".value"),
    names_pattern = "^(Spring|Summer|Fall|Winter)_(.+)$"
  ) %>%
  dplyr::mutate(
    season       = factor(season, levels = c("Spring", "Summer", "Fall", "Winter")),
    region       = factor(region, levels = c("Study Region", "WEA")),
    target_label = factor(target_label, levels = c(
      "North Atlantic Right Whale",
      "Blue Whale",
      "Baleen Whales (guild)",
      "Harbour Porpoise",
      "Small Odontocetes (guild)"
    ))
  ) %>%
  dplyr::filter(!is.na(season)) %>%
  dplyr::filter(!grepl("Deep Diver", target_label, ignore.case = TRUE)) %>%
  # Mean records per detection-year: total records / years with detections
  # Zero where no detections recorded in that season x region
  dplyr::mutate(
    mean_per_year = ifelse(n_years > 0, round(n_records / n_years, 1), 0)
  ) %>%
  # Join all-cetacean seasonal totals and calculate % of total cetacean records
  left_join(all_cet_seasonal, by = c("region", "season")) %>%
  dplyr::mutate(
    pct_of_cet = ifelse(
      !is.na(total_all_cet) & total_all_cet > 0,
      round(100 * n_records / total_all_cet, 1),
      NA_real_
    )
  )
```

### Seasonal distribution of sightings: WEA vs study region

Sightings of priority cetacean species and guilds were dplyr::summarised by season
(spring: March--May; summer: June--August; fall: September--November; winter:
December--February) across the full study region and within WEA grid cells
(cells with $\geq$ 25% overlap with designated WEA polygons). WEA records
are a spatial subset of study region records.

```{r seasonal-table, echo=FALSE, message=FALSE, warning=FALSE}
# ---- Table: mean records per detection-year by species x season x region --------
# Mean records/year accounts for uneven detection frequency across groups and
# seasons. More interpretable than raw totals for a multi-species summary.
# Cell format: mean records/yr (n years with detections).

table_data <- seasonal_long %>%
  dplyr::mutate(
    cell_val = ifelse(
      n_years > 0,
      paste0(mean_per_year, " (", n_years, " yr)"),
      "0"
    )
  ) %>%
  select(target_label, region, season, cell_val) %>%
  tidyr::pivot_wider(names_from = c(region, season), values_from = cell_val) %>%
  select(
    target_label,
    `Study Region_Spring`, `Study Region_Summer`,
    `Study Region_Fall`,   `Study Region_Winter`,
    `WEA_Spring`,          `WEA_Summer`,
    `WEA_Fall`,            `WEA_Winter`
  ) %>%
  rename("Species / Group" = target_label)

kable(
  table_data,
  caption = "Mean records per detection-year by season for priority cetacean
             species and guilds. Values show mean annual records (number of
             years with detections in parentheses); zero indicates no records
             in that season and region. Study Region covers the full analysis
             area; WEA values are a spatial subset (grid cells with >= 25%
             overlap with designated WEA polygons).",
  col.names = c(
    "Species / Group",
    "Spring", "Summer", "Fall", "Winter",
    "Spring", "Summer", "Fall", "Winter"
  ),
  align    = c("l", rep("c", 8)),
  booktabs = TRUE
) %>%
  kable_styling(
    bootstrap_options = c("striped", "condensed"),
    full_width        = TRUE,
    font_size         = 11
  ) %>%
  add_header_above(c(
    " "            = 1,
    "Study Region" = 4,
    "WEA"          = 4
  )) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, bold = TRUE, width = "18em")
```

```{r seasonal-plot, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=6, fig.cap="Seasonal sightings of priority cetacean species and guilds expressed as a percentage of total cetacean records in each region and season. Study Region (blue) covers the full analysis area; WEA (orange) is a spatial subset. Normalising to total cetacean figs/Effort_maps controls for the strong summer dominance in observer coverage, allowing seasonal patterns to be compared across groups. Seasons with zero records are omitted from bars."}
# ---- Bar chart: % of total cetacean records by season x species x region --------
# Normalising to total cetacean records per region x season controls for uneven
# observer figs/Effort_maps, making winter and spring detections visible relative to summer.

plot_data <- seasonal_long %>%
  dplyr::filter(n_records > 0, !is.na(pct_of_cet))

ggplot(plot_data, aes(x = season, y = pct_of_cet, fill = region)) +
  geom_col(
    position = position_dodge(width = 0.75),
    width    = 0.65,
    alpha    = 0.9
  ) +
  geom_text(
    aes(label = paste0(pct_of_cet, "%")),
    position = position_dodge(width = 0.75),
    vjust    = -0.4,
    size     = 2.6,
    color    = "grey30"
  ) +
  scale_fill_manual(
    values = c("Study Region" = "#6baed6", "WEA" = "#fd8d3c"),
    name   = NULL
  ) +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0.18))
  ) +
  facet_wrap(~ target_label, scales = "free_y", ncol = 3) +
  labs(
    x       = NULL,
    y       = "% of total cetacean records",
    caption = paste0(
      "Percentage = (species group records) / (all cetacean records) per region x season. ",
      "WEA = grid cells with >= 25% overlap with designated WEA polygons (25 km grid)."
    )
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.text         = element_text(size = 9, face = "bold"),
    legend.position    = "top",
    axis.text.x        = element_text(angle = 35, hjust = 1),
    panel.grid.major.x = element_blank(),
    plot.caption       = element_text(size = 8, color = "grey40", hjust = 0)
  )
```



---

# Script Structure & Logic Flow {#script-structure}

<details>
<summary>Click to expand — overview of all scripts and their section structure</summary>

**`setup/02_build_study_area.R`**
- One-time setup script; rebuild only if source spatial boundaries change.
- Load EEZ, NAFO, coastline, WEA shapefiles
- dplyr::filter NAFO to divisions 4V and 4W; union to single polygon
- Clip to WEA regional assessment study area

**`01_clean_aerial_sightings.R`**
- Read raw aerial survey data from `input/raw_data/aerial/`
- Standardize fields, filter marine mammals, deduplicate aerial observations
- Save processed aerial sightings to `input/processed_data/aerial/`

**`02_prepare_combined_sightings.R`**
- Configuration: file paths, spatial CRS, 1 km cell size
- Helper functions: WSDB datetime parser, HHMM time parser, grid cell ID creator
- Read reference data (WSDB species codes, aerial→WSDB species lookup)
- Read & standardize aerial data; read & standardize WSDB data; combine datasets
- Create 1 km spatial grid cells (UTM Zone 20N)
- Deduplication: 1 record per species per day per 1 km cell
- Deduplication summary statistics
- Save combined and deduplicated outputs + **`write_csv()` to `output/data/`**

**`03_plot_combined_sightings.R`**
- Load standard spatial layers via `helpers/helper_spatial.R`; read combined sightings CSV
- Define bounding box, create point shapefile, clip to study area
- dplyr::filter: cetacean codes only, not on land, ≥50 records
- Loop: generate per-species sightings maps → `ggsave`
- Facet map of all species → **`ggsave("output/figs/Map_Species...")`**

**`04_summarize_wsdb_effort.R`**
- Setup: packages, configuration (grid size, year range, target species, palette)
- Data loading & cleaning; grid cell calculation (25 km UTM cells)
- Load spatial layers; spatial tagging
- Aggregate: all cetacean records per grid cell → `cell_all`
- Confidence index: coverage quantiles × temporal consistency
- Coverage map → **`ggsave("output/figs/figs/Effort_maps/p_all...")`**
- Confidence map → **`ggsave("output/figs/figs/Effort_maps/p_conf...")`**
- Target species mapping; individual species maps

**`05_summarize_wsdb_confidence.R`**
- Requires `cell_all` and `targets` from `04_summarize_wsdb_effort.R`
- Confidence metrics function + target-specific metrics function
- Compile `overall_conf_summary`, `target_conf_summary`
- Save CSVs + visualisations (confidence comparison bar chart, target heatmap)
- **`write_csv()` to `output/data/`**

**`06_summarize_wsdb_seasonality.R`**
- Requires `dt` and `cell_m` from `04_summarize_wsdb_effort.R`
- Classify records by season; aggregate to seasonal grid summaries
- Seasonal faceted map → **`ggsave("output/figs/...")`**
- Seasonal bar chart → **`ggsave("output/figs/...")`**

**`07_summarize_pam_detections.R`**
- Read baleen presence CSV; assign seasons; compute seasonal detection summaries
- Plots p1–p5 → **`ggsave("output/figs/Effort_maps/PAM/...")`**
- Site distribution map by season → **`ggsave("output/figs/Effort_maps/PAM/...")`**
- PAM station acoustic coverage scenarios → **`ggsave("output/figs/Effort_maps/PAM/...")`**
- **`write_csv()` to `output/data/`**

**`08_run_kde_models.R`**
- Configuration, helper functions, KDE pipeline
- Output: rasters to `output/tif/`, shapefiles to `output/shapes/q90/`, plots to `output/figs/`

**`09_create_kde_contours.R`** — Batch multi-level contour shapefiles

**`10_compare_kde_pam_vs_sightings.R`**
- PAM vs sightings comparison maps → `output/figs/compare_sightings_pam/`

**`11_compare_kde_legacy.R`**
- Legacy PAM vs sightings comparison workflow retained for reference.

**`Studyareaclip_cetaceanfamily.r`** — Clips to study area, classifies by family

</details>

---

# Appendix A {#appendix-a}

```{r appendix-a-table-1}
# ---- Appendix A Table 1: WEA documented species ----
# Expected file: output/data/appendix_a_table1_wea_documented_species.csv

df_appendix_a_table1 <- read_csv_safe(csv_appendix_a_table1)

if (!is.null(df_appendix_a_table1)) {
  knitr::kable(
    df_appendix_a_table1,
    caption = paste(
      "Table 1 Appendix A. Species sighted within Wind Energy Areas (WEA).",
      "Established species are considered those with a minimum of 2 sightings records",
      "and 2 years of sightings within 10 km of any WEA boundary.",
      "FB = French Bank, MB = Middle Bank, SB = Sydney Bight, SIB = Sable Island Bank.",
      "Non-specific categories, Unidentified Whale and Dolphin/Porpoise (NS), are omitted from established species lists because they occurred as established groups in every WEA."
    ),
    align = c("l", "r", "r", "l")
  ) %>%
    kableExtra::kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = TRUE
    )
}
```

```{r appendix-a-table-2}
# ---- Appendix A Table 2: WEA comparability summary ----
# Expected file: output/data/appendix_a_table2_wea_comparability.csv

df_appendix_a_table2 <- read_csv_safe(csv_appendix_a_table2)

if (!is.null(df_appendix_a_table2)) {
  knitr::kable(
    df_appendix_a_table2,
    caption = paste(
      "Table 2 Appendix A. Comparability summary of cetacean data by WEA.",
      "Visual coverage summarizes all cetacean sighting records within 10 km of each WEA boundary,",
      "including WSDB and aerial survey records, year range, winter sighting presence,",
      "and the count of established species.",
      "Established species had at least 2 sighting records in at least 2 years within 10 km of the WEA boundary.",
      "Limited species were detected within 10 km of the WEA boundary but did not meet the established threshold.",
      "Non-specific categories, Unidentified Whale and Dolphin/Porpoise (NS), are omitted from established and limited species lists/counts.",
      "WEA = Wind Energy Area; WSDB = Whale Sightings Database."
    ),
    align = c("l", "r", "l", "l", "l")
  ) %>%
    kableExtra::kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = TRUE
    )
}
```

## Results Section 3.2 Copy-Ready Text

```{r appendix-a-comparability-text, results='asis'}
# ---- Copy-ready narrative for Word report Results section 3.2 ----
# The text mirrors the current Word draft section and updates the values from
# reproducible CSVs generated by 02_prepare_combined_sightings.R,
# 04_summarize_wsdb_effort.R, and 07_summarize_pam_detections.R.

sightings_summary <- read_csv_safe(csv_sightings_summary)
sightings_by_source <- read_csv_safe(csv_sightings_by_source)
sightings_by_species <- read_csv_safe(csv_sightings_by_species)
table1_established <- read_csv_safe(csv_appendix_a_table1)
table2_stats <- read_csv_safe(csv_appendix_a_table2_stats)
wea_species_status <- read_csv_safe(csv_appendix_a_species_status)
wea_10km_species <- read_csv_safe(csv_wea_10km_species)
wea_10km_area <- read_csv_safe(csv_wea_10km_area)
pam_wea_distance <- read_csv_safe(csv_pam_wea_distance)

# External unpublished DFO note, not part of the 2010-2024 analysis dataset.
killer_2026_wea_sightings <- 2

format_list <- function(x) {
  x <- x[!is.na(x) & x != ""]
  if (length(x) == 0) return("")
  if (length(x) == 1) return(x)
  if (length(x) == 2) return(paste(x, collapse = " and "))
  paste0(paste(x[-length(x)], collapse = ", "), ", and ", x[length(x)])
}

get_source_n <- function(source_name) {
  sightings_by_source %>%
    dplyr::filter(source == source_name) %>%
    dplyr::pull(n_records) %>%
    dplyr::first()
}

get_wea <- function(wea_name) {
  table2_stats %>%
    dplyr::filter(WEA == wea_name) %>%
    dplyr::slice(1)
}

get_species_10km <- function(species_name) {
  out <- wea_10km_species %>%
    dplyr::filter(species == species_name) %>%
    dplyr::slice(1)

  if (nrow(out) == 0) {
    return(tibble::tibble(
      species = species_name,
      n_study_area = 0,
      n_wea_10km = 0,
      pct_wea_10km = 0,
      years_study_area = 0,
      years_wea_10km = 0
    ))
  }

  out
}

get_area_row <- function(region_name) {
  wea_10km_area %>%
    dplyr::filter(region == region_name) %>%
    dplyr::slice(1)
}

if (!is.null(sightings_summary) &&
    !is.null(sightings_by_source) &&
    !is.null(sightings_by_species) &&
    !is.null(table1_established) &&
    !is.null(table2_stats) &&
    !is.null(wea_species_status) &&
    !is.null(wea_10km_species) &&
    !is.null(wea_10km_area) &&
    !is.null(pam_wea_distance)) {

  study <- sightings_summary[1, ]
  ns_species <- sightings_by_species %>%
    dplyr::filter(stringr::str_detect(common_name, "\\(NS\\)|Unidentified")) %>%
    dplyr::pull(common_name) %>%
    sort()

  species_status_summary <- wea_species_status %>%
    dplyr::summarise(
      identified_species_n = dplyr::n_distinct(species[include_in_species_lists]),
      ns_group_n = dplyr::n_distinct(species[!include_in_species_lists])
    )

  total_wea_records <- sum(table2_stats$total_records, na.rm = TRUE)
  established_species_n <- nrow(table1_established)

  sydney <- get_wea("Sydney Bight")
  french <- get_wea("French Bank")
  middle <- get_wea("Middle Bank")
  sable <- get_wea("Sable Island Bank")

  study_area_row <- get_area_row("Study area")
  wea_buffer_area <- get_area_row("WEA 10 km buffer")

  minke <- get_species_10km("Minke Whale")
  humpback <- get_species_10km("Humpback Whale")
  aws_dolphin <- get_species_10km("Atlantic White-Sided Dolphin")
  fin <- get_species_10km("Fin Whale")
  sei <- get_species_10km("Sei Whale")
  common_dolphin <- get_species_10km("Common Dolphin")
  pilot <- get_species_10km("Long-Finned Pilot Whale")
  killer <- get_species_10km("Killer Whale")
  narw <- get_species_10km("North Atlantic Right Whale")

  # Build top-4 list after Minke, sorted descending by % within 10 km of WEA.
  # Excludes Minke (already the lead sentence), Pilot, Killer, NARW, NS groups,
  # and Common Dolphin (mentioned separately in the next sentence).
  notable_candidates <- dplyr::bind_rows(
    humpback, aws_dolphin, fin, sei
  ) %>%
    dplyr::arrange(dplyr::desc(pct_wea_10km))

  # Format each as "Species name (XX%)"
  notable_labels <- paste0(
    notable_candidates$species, " (",
    round(notable_candidates$pct_wea_10km), "%)"
  )
  # Collapse into readable list: "A (x%), B (x%), C (x%), and D (x%)"
  notable_text <- if (length(notable_labels) == 1) {
    notable_labels
  } else {
    paste0(
      paste(notable_labels[-length(notable_labels)], collapse = ", "),
      ", and ", notable_labels[length(notable_labels)]
    )
  }
  unidentified_whale <- get_species_10km("Unidentified Whale")
  dolphin_ns <- get_species_10km("Dolphin/Porpoise (NS)")

  pam_min <- pam_wea_distance %>%
    dplyr::arrange(nearest_station_km) %>%
    dplyr::slice(1)
  pam_max <- pam_wea_distance %>%
    dplyr::arrange(dplyr::desc(nearest_station_km)) %>%
    dplyr::slice(1)
  pam_within_10km <- sum(pam_wea_distance$stations_within_10km, na.rm = TRUE)
  french_middle_year_text <- if (french$years_count == middle$years_count) {
    paste0("each across ", french$years_count, " years")
  } else {
    paste0(
      "across ", french$years_count, " and ", middle$years_count,
      " years, respectively"
    )
  }

  # Helper: spell out small integers as words (for natural prose)
  num_to_words <- function(n) {
    words <- c("zero","one","two","three","four","five","six","seven","eight",
               "nine","ten","eleven","twelve","thirteen","fourteen","fifteen",
               "sixteen","seventeen","eighteen","nineteen","twenty")
    if (n >= 0 && n <= 20) words[n + 1] else as.character(n)
  }

  # French + Middle established taxa: use max if equal, otherwise list both
  french_middle_established <- if (french$species_n_established == middle$species_n_established) {
    paste0(french$species_n_established, " established taxa overall")
  } else {
    paste0(french$species_n_established, " and ", middle$species_n_established,
           " established species respectively")
  }

  cat(
    paste0(
      "Across the study area, the cetacean sightings dataset compiled for this report consisted of ",
      scales::comma(study$n_records), " records between ",
      lubridate::year(study$date_min), " and ", lubridate::year(study$date_max),
      ", including ", scales::comma(get_source_n("wsdb")),
      " opportunistic records from the WSDB and ",
      scales::comma(get_source_n("aerial")),
      " aerial survey records (Figure 2A). Records reflected ",
      study$n_species, " unique taxa or taxonomic groupings of cetaceans, including ",
      study$n_species_ns, " non-specified cetacean categories (i.e., ",
      format_list(ns_species), "; Appendix C). Sightings were strongly summer-weighted and provided limited winter coverage across most of the study area (Figure 3), constraining inference on year-round occurrence.\n\n",

      "Drawing on sightings records from ", lubridate::year(study$date_min),
      " to ", lubridate::year(study$date_max),
      ", visual detections within a 10km buffer of the four WEA boundaries documented a diverse cetacean assemblage with ",
      species_status_summary$identified_species_n, " identified species and ",
      species_status_summary$ns_group_n, " non-specific (NS) groupings. Within a 10km buffer there were ",
      scales::comma(total_wea_records),
      " records across all four WEAs combined. Established species, those with a minimum of 2 records and 2 years of detections within 10 km of WEA boundaries, were concentrated across a smaller subset of ",
      num_to_words(established_species_n), " identified species (Table 1).\n\n",

      "Coverage varied considerably across the four WEAs. Sydney Bight was the best-characterised area, with ",
      scales::comma(sydney$total_records), " cetacean records in ",
      sydney$years_count, " years, overlapping with aerial survey effort, winter visual detections, and ",
      sydney$species_n_established, " established species. French Bank and Middle Bank each contributed ",
      scales::comma(french$total_records), " and ",
      scales::comma(middle$total_records), " records respectively, each across ",
      french$years_count, " years, with ", french_middle_established,
      ", though neither area had winter sighting records. Sable Island Bank had the fewest cetacean sightings records (n = ",
      scales::comma(sable$total_records), ") despite being the largest WEA by area (",
      scales::comma(sable$Area_km2), " km²), with ",
      sable$species_n_established,
      " established species and both aerial and winter sightings.\n\n",

      "PAM coverage was contextual for all four WEAs, as ",
      ifelse(pam_within_10km == 0, "no stations fell", paste(pam_within_10km, "stations fell")),
      " within 10 km of any WEA boundary, and the nearest stations ranged from approximately ",
      round(pam_min$nearest_station_km), " km (", pam_min$WEA, ") to ",
      round(pam_max$nearest_station_km), " km (", pam_max$WEA,
      ") outside the WEAs, providing acoustic context for adjacent waters but not direct within-WEA characterisation (Table 2).\n\n",

      "Relative to the full study area (", scales::comma(round(study_area_row$area_km2)),
      " km²), the 10 km WEA buffer footprint (",
      scales::comma(round(wea_buffer_area$area_km2)),
      " km²) represented ", round(wea_buffer_area$pct_study_area),
      "% of the total area yet still captured a notable share of detections for some taxa. Minke whale showed the strongest concentration among frequently recorded species, with approximately ",
      round(minke$pct_wea_10km),
      "% of all study-area sightings occurring within 10 km of WEA boundaries. ",
      notable_text,
      " also showed notable proportions. Common dolphin contributed approximately ",
      round(common_dolphin$pct_wea_10km),
      "% of study-area records within 10 km of WEA boundaries, while long-finned pilot whale, despite being among the most frequently recorded species in the broader study area (",
      scales::comma(pilot$n_study_area),
      " study-area records), contributed only about ",
      round(pilot$pct_wea_10km),
      "% within 10 km of WEA boundaries, indicating uneven distribution across the study area.\n\n",

      "Killer whale and North Atlantic right whale had ",
      scales::comma(killer$n_wea_10km), " and ",
      scales::comma(narw$n_wea_10km),
      " records, respectively, within 10 km of WEA boundaries and cannot be considered established at this time. Killer whale records are generally rare in the region, with only ",
      scales::comma(killer$n_study_area),
      " study-area records within the current dataset, although dedicated marine mammal surveys in 2026 documented ",
      scales::comma(killer_2026_wea_sightings),
      " sightings inside the WEAs (unpublished data, DFO). North Atlantic right whale sightings records in the study area are also very sparse (n = ",
      scales::comma(narw$n_study_area),
      "), making it difficult to draw conclusions about WEA-level occurrence or habitat use from the current dataset.\n\n",

      "Records of non-specified cetaceans are informative for broad occurrence screening and suggest there are more cetaceans (n = ",
      scales::comma(unidentified_whale$n_wea_10km),
      ") and dolphins or porpoises (n = ",
      scales::comma(dolphin_ns$n_wea_10km),
      ") within 10 km of WEA boundaries than could be identified to the species level.\n\n"
    )
  )
}
```

## Results Section 3.3 Copy-Ready Text

```{r section-3-3-seasonal, results='asis'}
# ---- Copy-ready narrative for Word report Results section 3.3 ----
# Reads seasonal_records_summary.csv saved by 06_summarize_wsdb_seasonality.R.
# Numbers reflect the same dataset and filters used by that script (cell_m,
# year >= 2010, study area extent), so they will always match the season maps.
# Requires: output/data/seasonal_records_summary.csv

seas_csv <- here("output", "data", "seasonal_records_summary.csv")

if (file.exists(seas_csv)) {

  seas_summary <- readr::read_csv(seas_csv, show_col_types = FALSE) %>%
    dplyr::mutate(season = factor(season, levels = c("Spring", "Summer", "Fall", "Winter")))

  # Helper: pull a single value for a given season and column
  s <- function(season_name, col) {
    seas_summary[[col]][seas_summary$season == season_name]
  }

  cat(
    paste0(
      "As noted above, sightings records were strongly summer-weighted: summer accounted for ",
      s("Summer", "pct"), "% of all records (n = ",
      scales::comma(s("Summer", "n_records")),
      "), while winter represented only ",
      s("Winter", "pct"), "% (n = ",
      scales::comma(s("Winter", "n_records")),
      ") (Figure 3). Spring and fall contributed ",
      s("Spring", "pct"), "% (n = ",
      scales::comma(s("Spring", "n_records")), ") and ",
      s("Fall", "pct"), "% (n = ",
      scales::comma(s("Fall", "n_records")),
      "), respectively. ",
      "This seasonal imbalance was reflected spatially, with summer sightings distributed broadly across the study area and winter records sparse and patchy, particularly across offshore areas (Figure 3). ",
      "Consequently, the dataset provides stronger support for inference on occurrence during spring through fall than during winter. ",
      "Areas with few or no winter sightings should be interpreted cautiously due to limited observation coverage."
    )
  )

} else {
  cat("*Seasonal summary not available -- run 06_summarize_wsdb_seasonality.R (RUN_SUMMARIZE_SEASONALITY <- TRUE) first.*")
}
```

## Results Section 3.5 Copy-Ready Text

```{r section-3-5-confidence, results='asis'}
# ---- Copy-ready narrative for Word report Results section 3.5 ----
# Reads overall_conf_summary.csv (from 05_summarize_wsdb_confidence.R) and
# wea_10km_area_summary.csv (from 02_prepare_combined_sightings.R).
# Grid cell size is read from params_effort.csv.
# Spatial extent is expressed as % of study area (cells * cell_km2 / study_area_km2).
# Requires: output/data/overall_conf_summary.csv
#           output/data/wea_10km_area_summary.csv
#           output/data/params_effort.csv

conf_csv <- here("output", "data", "overall_conf_summary.csv")
area_csv <- here("output", "data", "wea_10km_area_summary.csv")
par_csv  <- here("output", "data", "params_effort.csv")

if (file.exists(conf_csv) && file.exists(area_csv) && file.exists(par_csv)) {

  conf  <- readr::read_csv(conf_csv, show_col_types = FALSE)
  areas <- readr::read_csv(area_csv, show_col_types = FALSE)
  pars  <- readr::read_csv(par_csv,  show_col_types = FALSE)

  # Grid cell area in km²
  grid_km      <- as.numeric(pars$Value[pars$Parameter == "grid_km"])
  cell_area_km2 <- grid_km^2

  # Study area km² (for converting cell counts to % of study area)
  study_area_km2 <- areas$area_km2[areas$region == "Study area"]
  wea_area_km2   <- areas$area_km2[areas$region == "WEA polygons"]

  # Pull study region and WEA rows
  sr  <- conf[conf$region == "Study Region", ]
  wea <- conf[conf$region == "WEA", ]

  # Approximate gridded area (cells × cell size; stated as approximate because
  # 25 km grid cells slightly overhang the study area boundary)
  sr_area_approx  <- round(sr$total_cells  * cell_area_km2 / 1000) * 1000  # round to nearest 1000
  wea_area_approx <- round(wea$total_cells * cell_area_km2 / 100)  * 100   # round to nearest 100

  cat(
    paste0(
      "Across the full study region, sightings were recorded in ",
      ifelse(sr$cells_with_data == sr$total_cells,
             "every grid cell (approximately ",
             paste0(sr$cells_with_data, " of ", sr$total_cells, " grid cells (~")),
      scales::comma(sr_area_approx), " km²), ",
      "with a mean of ", sr$mean_records_per_cell,
      " records per ", grid_km, " km cell (median ", sr$median_records_per_cell,
      ") and mean temporal coverage of ", sr$mean_years_per_cell,
      " years (median ", sr$median_years_per_cell, ") (Figure 5). ",
      "High confidence was reached in ", sr$pct_high_confidence,
      "% of cells; ", sr$pct_low_confidence, "% were classified as low confidence. ",

      "Within the WEA footprint, sightings were recorded in ",
      ifelse(wea$cells_with_data == wea$total_cells,
             "every grid cell",
             paste0(wea$cells_with_data, " of ", wea$total_cells, " grid cells")),
      ", ",
      "with a mean of ", wea$mean_records_per_cell,
      " records per ", grid_km, " km cell (median ", wea$median_records_per_cell,
      ") and mean temporal coverage of ", wea$mean_years_per_cell,
      " years (median ", wea$median_years_per_cell, "). ",
      "High confidence was reached in ", wea$pct_high_confidence,
      "% of WEA cells, while ", wea$pct_low_confidence, "% were classified as low confidence",
      ifelse(wea$pct_low_confidence > sr$pct_low_confidence,
             " -- a somewhat higher proportion than the study region overall, consistent with the offshore position of WEAs relative to primary survey effort.",
             " -- consistent with the study region overall."),

      " Confidence was highest for shelf baleen whales and small odontocetes, and lowest for conservation-priority, cryptic, and deep-diving taxa, as detailed in section 3.6."
    )
  )

} else {
  cat("*Confidence summary not available -- run 04_summarize_wsdb_effort.R and 05_summarize_wsdb_confidence.R first.*")
}
```

## Results Section 3.6 Copy-Ready Text — Baleen Whales

```{r section-3-6-baleen, results='asis'}
# ---- Copy-ready narrative for Word report Results section 3.6 (Baleen Whales) ----
# Reads target_conf_summary.csv produced by 05_summarize_wsdb_confidence.R.
# The "Shelf_baleen" target covers blue, fin, sei, humpback, minke, NARW, and
# unidentified baleen whales. Geographic/qualitative text is fixed; numbers
# (cells, records, mean, confidence %) are pulled from data.
# Requires: output/data/target_conf_summary.csv

tgt_csv <- here("output", "data", "target_conf_summary.csv")

if (file.exists(tgt_csv)) {

  tgt <- readr::read_csv(tgt_csv, show_col_types = FALSE)

  # Study Region row for shelf baleen
  b <- tgt %>%
    dplyr::filter(target == "Shelf_baleen", region == "Study Region") %>%
    dplyr::slice(1)

  cat(
    paste0(
      "The baleen whale group included blue, fin, sei, humpback, minke, and North Atlantic right whales, ",
      "as well as baleen whales that could not be identified to species. ",
      "Effort-normalized sightings showed a higher proportional representation of baleen whales ",
      "in the Cabot Strait near Sydney Bight and between Middle and French Banks, ",
      "where records spanned multiple years (Figure 6). ",
      "Low data cells appeared across the southern half of the study area, ",
      "overlapping French, as well as Sable Banks. ",
      "Confidence in sightings data on baleen whales was high, with records across ",
      round(b$cells_with_target / b$total_cells * 100), "% of the study area (",
      scales::comma(b$total_target_records), " total records, mean ",
      b$mean_target_per_cell, " records per 25 km² cell), with ",
      b$pct_moderate_plus, "% of cells reaching moderate or higher confidence and ",
      round(b$pct_no_evidence), "% showing no evidence."
    )
  )

} else {
  cat("*Target confidence summary not available -- run 04_summarize_wsdb_effort.R and 05_summarize_wsdb_confidence.R first.*")
}
```

## Results Section 3.6.2 Copy-Ready Text — Small Odontocetes and Dolphins

```{r section-3-6-2-odontocetes, results='asis'}
# ---- Copy-ready narrative for Word report Results section 3.6.2 ----
# Reads target_conf_summary.csv produced by 05_summarize_wsdb_confidence.R.
# "Small_odontocetes" target covers common dolphins, Atlantic white-sided dolphins,
# white-beaked dolphins, harbour porpoise, pilot whales, killer whales, and
# unidentified dolphins/small odontocetes.
# Geographic/qualitative text is fixed; numbers are pulled from data.
# Requires: output/data/target_conf_summary.csv

if (file.exists(tgt_csv)) {

  o <- tgt %>%
    dplyr::filter(target == "Small_odontocetes", region == "Study Region") %>%
    dplyr::slice(1)

  cat(
    paste0(
      "The small odontocete and dolphin group included common dolphins, Atlantic white-sided dolphins, ",
      "white-beaked dolphins, harbour porpoise, pilot whales, killer whales, and unidentified dolphins ",
      "or small odontocetes. ",
      "Sightings-based maps indicated a higher proportion of occurrence along coastal and Scotian slope areas, ",
      "including higher areas of concentration north of Sydney Bight and overlapping with French and Middle Banks (Figure 8-9). ",
      "Low data cells appeared across the study area, overlapping French, as well as Sable Banks (Figure 8). ",
      "Because there are no corresponding PAM analyses for odontocetes in this report, we have limited ",
      "understanding of seasonal persistence across the study area. ",
      "Small odontocetes showed the strongest coverage overall, with records across ",
      round(o$cells_with_target / o$total_cells * 100), "% of the study area (",
      scales::comma(o$total_target_records), " total records, mean ",
      o$mean_target_per_cell, " records per 25 km² cell), with ",
      o$pct_moderate_plus, "% of study region cells at moderate or higher confidence and ",
      round(o$pct_no_evidence), "% lacking any records."
    )
  )

} else {
  cat("*Target confidence summary not available -- run 04_summarize_wsdb_effort.R and 05_summarize_wsdb_confidence.R first.*")
}
```

## Results — Rare Taxa Copy-Ready Text

```{r section-rare-taxa, results='asis'}
# ---- Copy-ready narrative for rare taxa paragraph ----
# Reads sightings_by_species.csv (produced by 02_prepare_combined_sightings.R).
# Species are sorted by record count descending and listed with inline n values.
# Requires: output/data/sightings_by_species.csv

sp_csv <- here("output", "data", "sightings_by_species.csv")

if (file.exists(sp_csv)) {

  sp <- readr::read_csv(sp_csv, show_col_types = FALSE)

  # Helper: pull n for a named species (returns 0 if not found)
  sp_n <- function(name) {
    val <- sp$n[sp$common_name == name]
    if (length(val) == 0) 0L else as.integer(val[1])
  }

  # Rare taxa list -- add or remove species here as needed
  rare <- tibble::tribble(
    ~label,                  ~key,
    "striped dolphin",       "Striped Dolphin",
    "white-beaked dolphin",  "White-Beaked Dolphin",
    "sei whale",             "Sei Whale",
    "killer whale",          "Killer Whale",
    "beluga whale",          "Beluga Whale",
    "false killer whale",    "False Killer Whale"
  ) %>%
    dplyr::mutate(n = purrr::map_int(key, sp_n)) %>%
    dplyr::arrange(dplyr::desc(n))

  # Build "Species (n = X), Species (n = X), ..., and Species (n = X)" list
  entries <- paste0(rare$label, " (n = ", scales::comma(rare$n), ")")
  rare_list <- if (length(entries) == 1) {
    entries
  } else {
    paste0(paste(entries[-length(entries)], collapse = ", "),
           ", and ", entries[length(entries)])
  }

  cat(
    paste0(
      "Records of ", rare_list,
      " were too sparse to support standalone spatial interpretation but are noted as ",
      "relevant evidence of occurrence and species diversity for the broader Scotian Shelf study area."
    )
  )

} else {
  cat("*Species summary not available -- run 02_prepare_combined_sightings.R first.*")
}
```

---

*Report generated with `rmarkdown::render()`. To re-knit:*

```r
rmarkdown::render("KDE_PAM_cetaceans_report.Rmd")
```

*To regenerate CSV summary files (one-time, then report reads from disk):*

```r
# Add write_csv() calls to each script as noted in chunk comments above,
# then source() each script once. The report will read CSVs on all future knits.
```
